---
title: "Plotting"
output: rmarkdown::html_vignette
fig_height: 5
fig_width: 5
vignette: >
  %\VignetteIndexEntry{Plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(arfpam)
```

## Table of contents {#TOC}

0. [Preamble](#section0)
1. [Example histogram](#section1)
2. [Example dot plot](#section2)
3. [Example forest plot](#section3)
4. [Example line plot](#section4)
5. [Plotting over groups](#section5)
6. [Example correlation heatmap](#section6)

## 0) Preamble {#section0}

Before providing examples of the various plotting functions, we define two useful functions to quickly create descriptive summaries of our data of interest:
```{r}
quick_mean_and_ci <- function(x) {
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  se <- s/sqrt(n)
  
  out <- c( m, m + se*qt( c( .025, .975 ), n - 1  ) )
  names(out) <- c( 'M', 'LB', 'UB' )
  
  return(out)
}

summary_by_groups <- function(dtf, x, groupings) {
  
  lst_groups <- dtf[, groupings]
  if (length(groupings) == 1) lst_groups <- list(lst_groups)
  names(lst_groups) <- groupings
  
  sm <- aggregate(
    dtf[[x]],
    lst_groups,
    quick_mean_and_ci
  )
  sm <- cbind(
      sm[, seq_along(groupings)],
      sm$x
    )
  
  if ( length(groupings) == 1 ) {
    colnames(sm)[1] <- groupings
  }
  if ( is.matrix(sm) ) sm <- data.frame(sm)
  
  return(sm)
}
```

## 1) Example histogram {#section1}

One can quickly plot a histogram for a set of values via the 'plot_histogram' function, which uses by default the Freedman-Diaconis rule for determining bin size (which works somewhat better then base R's default using Sturge's rule):

```{r}
# Simulate data from normal distribution
x <- rnorm(100)
plot_histogram(x, main = 'Normal distribution', new = FALSE )

# Simulate data from log-normal distribution
y <- exp( rnorm(100) )
plot_histogram(y, main = 'Log-normal distribution', new = FALSE )
```

[Table of contents](#TOC)

## 2) Example dot plot {#section2}

A common figure type in psychology is a plot of a measure central tendency and its variation (e.g., means and 95% confidence intervals), shown over a grouping factor. Such a figure can be created quickly using the 'draw_dots' function:

```{r}
# Example data examining effect of diet on early growth of chicks
data("ChickWeight")

# Create descriptive summary by Diet for final day
lgc_final <- ChickWeight$Time == 21
dtf_obs <- summary_by_groups( ChickWeight[lgc_final, ], 'weight', 'Diet' )
dtf_obs$X <- 1:nrow( dtf_obs )
# See the package 'dplyr' for a concise way to create these summaries

# Plot means and 95% confidence intervals for weight

# First create blank plot
xl <- c( .5, 4.5 )
yl <- c( 120, 340 )
blank_plot( xl, yl )

# Add estimates and error bars
draw_dots( dtf_obs, columns = c( 'X', 'M', 'LB', 'UB' ) )

# Add borders, labels, and axes
draw_borders_and_labels(
  xl, yl, labels = c( 'Diet', 'Weight at 21 days (gm)' )
)
add_axes( seq( yl[1], yl[2], 40 ), side = 2, line = -1.25, cex = 1 )
add_axes( 1:4, 'Diet ' %p% 1:4, side = 1, line = -1.25, cex = 1 )
```

[Table of contents](#TOC)

## 3) Example forest plot {#section3}

A useful variant of figures summarizing estimates and error bars is the forest plot, used commonly to summarize the results of a meta-analysis. In its most basic form, a forest plot reports a set of estimates and associated error bars for different variables:

```{r}
generating_means <- runif( 6, -1, 1 )
x <- MASS::mvrnorm( 100, generating_means, diag(6) )
dtf <- apply( x, 2, quick_mean_and_ci ) |> t()
dtf <- data.frame(dtf)

plot_forest(
  dtf, xlim = c(-1.5, 1.5), 
  labels_y = 'Column ' %p% 1:6, 
  labels_x = seq( -1.5, 1.5, .5 ), 
  title_x = 'Estimated mean', 
  new = FALSE
)
points( generating_means, 6:1, pch = 21, bg = 'white', cex = 1.25 )

legend(
  -1.5, 7, 'Generating mean', pch = 21, bty = 'n', horiz = TRUE, xpd = NA
)
```

[Table of contents](#TOC)

## 4) Example line plot {#section4}

Another common figure in psychology is a line plot displaying change in a variable over time. We can quickly create such a figure using the 'draw_line' function:

```{r}
# Create descriptive summary by time collapsing over diet
dtf_obs <- summary_by_groups(
  ChickWeight, 'weight', 'Time'
)

# Plot means and 95% confidence intervals for weight

# First create blank plot
xl <- c( -.5, 21.5 )
yl <- c( 0, 250 )
blank_plot( xl, yl )

# Add estimates and error bars
draw_lines( dtf_obs, columns = c( 'Time', 'M', 'LB', 'UB' ), col.eb = 'grey' )

# Add borders, labels, and axes
draw_borders_and_labels(
  xl, yl, labels = c( 'Day', 'Weight (gm)' )
)
add_axes( seq( yl[1], yl[2], 50 ), side = 2, line = -1.25, cex = 1 )
add_axes( seq( 0, 20, 5 ), side = 1, line = -1.25, cex = 1 )
```

[Table of contents](#TOC)

## 5) Plotting over groups {#section5}

Often we will need to plot multiple trajectories over time for separate groups. The 'draw_by_groups' function streamlines the process of plotting separate lines by different groups.

```{r}
# Create descriptive summary across both time and diet
dtf_obs <- summary_by_groups(ChickWeight, 'weight', c( 'Diet', 'Time' ) )
dtf_obs$X <- dtf_obs$Time + 
  replace_cases( dtf_obs$Diet, 1:4, c( -.6, -.2, .2, .6 ) )
dtf_obs$col <- 
  replace_cases( dtf_obs$Diet, 1:4, palettes( index = 1:4 ) )
# See the package 'dplyr' for a concise way to create these summaries

# Plot means and 95% confidence intervals for weight

# First create blank plot
xl <- c( -1, 22 )
yl <- c( 0, 350 )
blank_plot( xl, yl )

draw_by_group(
  dtf_obs, 'Diet', 1:4,
  draw_fun = draw_lines,
  columns = c( 'X', 'M', 'LB', 'UB' ),
  arrow = TRUE,
  pch = 21, 
  aes = c( col = 'col', col.eb = 'col', bg = 'col' )
)

# Add borders, labels, and axes
draw_borders_and_labels(
  xl, yl, labels = c( 'Day', 'Weight (gm)' )
)
add_axes( seq( yl[1], yl[2], 50 ), side = 2, line = -1.25, cex = 1 )
add_axes( seq( 0, 20, 5 ), side = 1, line = -1.25, cex = 1 )
legend(
  'topleft', 'Diet ' %p% 1:4, fill = palettes( 1:4 ), bty = 'n'
)
```

[Table of contents](#TOC)

## 6) Example correlation heatmap {#section6}

The function 'plot_correlation_heatmap' is a quick way to create a figure summarizing the set of correlations over multiple variables along with useful information on the magnitude and statistical significant of each relationship:

```{r}
# Simulate data from a multivariate normal with correlated values
Sigma <- rbind(
  c( 1.0, 0.2, 0.5 ),
  c( 0.2, 1.0, 0.1 ),
  c( 0.5, 0.1, 1.0 )
)
x <- MASS::mvrnorm( 100, rep( 0, 3 ), Sigma = Sigma )
colnames(x) <- 'V' %p% 1:3
x <- data.frame(x)

plot_correlation_heatmap(x, new = FALSE)
```

[Table of contents](#TOC)

